results-draft

### 2024-03-02 archive/reults-draft.Rmd
results draft from google docs

---
title: "The value of early warning systems for pandemic response"
format:
  html:
    self-contained: true
execute: 
  echo: true
  warning: false
  error: false
author: "Pedro Nascimento de Lima"
---

@@

# Population
The model simulates disease percolating through six interconnected jurisdictions that independently make nonpharmaceutical decisions. Individuals can travel and work across jurisdiction lines, and the six jurisdictions each have a population of 10^6 individuals. Mortality parameters were chosen to reflect mortality rates observed in the United States during the first year of the COVID-19 pandemic, hence our results can be interpreted as expected outcomes for a cluster of US counties.

@@

# Calibration: Add as a footnote to tau.
We use parameter estimates from the literature whenever possible (Supplementary Table 1). Parameters that are unobserved in the data, notably tau and. R0 is computed based on disease progression rates. We calibrate only the $tau$ parameter, which is highly heterogeneous and context-dependent (cite Laura), to match the number of deaths observed during the first year in the United States. We find that $tau = 0.137$. We explore the implication of a wider range of policy effectiveness in sensitivity analyses, and explore this particular parameter in more details. Surprisingly, the policy effectiveness

@@

<!--
Points from this section for discussion:
* **EWS provide meaningful value under base-case assumptions.** Net-monetary benefit from surveillance system is on the order of 1,000 dollars per person over a one-year time-frame. This outweighs any available estimate of costs.
* **Base-case results align with observed COVID-19 data.** US obseved on the order of 160 deaths per 100,000 people during the first year of the pandemic, and many states.
-->

## Value of EWS system depends on epidemiological and societal parameters

@@

A pathogen that is 50% more transmissible than wild-type SARS-Cov-2 would pose a greater challenge for pandemic response, causing `r pull_estimate(r$table_1_long_all, Scenario == "1.5x transmissible", variable == "deaths_per_100k", NMB_comparator == T)` deaths per 100,000 people even if mitigated with the same NPI policy considered in our base-case scenario (Supplementary Table 2). Such a pandemic would cause a substantial societal cost of `r pull_estimate(r$table_1_long_all, Scenario == "1.5x transmissible", variable == "C", NMB_comparator == T)` dollars per person. In those conditions, the EWS would be particularly valuable, reducing the death toll of the pandemic to `r pull_estimate(r$table_1_long_all, Scenario == "1.5x transmissible", variable == "deaths_per_100k", NMB_comparator == F)` per 100,000 people, and would be projected to provide a NMB of `r pull_estimate(r$table_1_long_all, Scenario == "1.5x transmissible", variable == "NMB", NMB_comparator == F)` dollars per person. Conversely, a pathogen that is 50% less transmissible than SARS-Cov-2 would result in a lower death toll of `r pull_estimate(r$table_1_long_all, Scenario == "0.5x transmissible", variable == "deaths_per_100k", NMB_comparator == T)` without the EWS system, reducing the net-monetary benefit of the system to `r pull_estimate(r$table_1_long_all, Scenario == "0.5x transmissible", variable == "NMB", NMB_comparator == F)` dollars per person. The value of the EWS system also depends on the infection fatality rate of the pathogen, but the net-benefit of the EWS would be positive even for a pathogen that is half as deadly as SARS-Cov-2. All else equal, the EWS would still yield an NMB of `r pull_estimate(r$table_1_long_all, Scenario == "0.5x deadly", variable == "NMB", NMB_comparator == F)` dollars per person in a pandemic caused by a pathogen that is 50% less deadly than SARS-Cov-2, and would yield `r pull_estimate(r$table_1_long_all, Scenario == "1.5x deadly", variable == "NMB", NMB_comparator == F)` dollars per person in responding to a pathogen that is 50% more deadly than COVID-19.

@@

Longer decision and implementation lags have an overall negative effect on the total costs of the pandemic, longer decision lags also would *increase* the net benefit of EWS systems for pandemic response. Our base-case scenario assume that decision-makers take one week to decide and implement non-pharmaceutical interventions. If instead policymakers took two weeks to introduce and implement interventions, the total pandemic cost increases from `r pull_estimate(r$table_1_long_all, Scenario == "Base-case", variable == "C", NMB_comparator == T)` to `r pull_estimate(r$table_1_long_all, Scenario == "2-week decision lag", variable == "C", NMB_comparator == T)` dollars per person, assuming conventional surveillance. The early warning system effectively helps to roll-back decision lags, reducing total costs to `r pull_estimate(r$table_1_long_all, Scenario == "2-week decision lag", variable == "C", NMB_comparator == F)` dollars per person, yielding a NMB of `r pull_estimate(r$table_1_long_all, Scenario == "2-week decision lag", variable == "NMB", NMB_comparator == F)` relative to conventional surveillance. Conversely, in contexts where NPI implementation is swift (i.e, decisions are made within 3 days), the NMB of EWS would be lower but still positive at `r pull_estimate(r$table_1_long_all, Scenario == "3-day decision lag", variable == "NMB", NMB_comparator == F)` dollars per person.

Policy stringency (i.e., how stringent the case thresholds used to introduce or remove interventions) has a similar effect on the value of EWS relative to decision lags - the EWS provides the most value under the least desirable conditions for disease prevention. In a context where decision-makers are 50% more stringent than in our base-case assumptions (i.e., their case thresholds are half of its base-case), society is able to avert a substantial number of deaths, reducing mortality from `r pull_estimate(r$table_1_long_all, Scenario == "Base-case", variable == "deaths_per_100k", NMB_comparator == T)` to `r pull_estimate(r$table_1_long_all, Scenario == "1.5x stringent", variable == "deaths_per_100k", NMB_comparator == T)` deaths per 100,000 population without the EWS. In that scenario, the EWS yields `r pull_estimate(r$table_1_long_all, Scenario == "1.5x stringent", variable == "NMB", NMB_comparator == F)` dollars per person of net benefit since most of the deaths would already be prevented through more stringent NPIs. Conversely, if policy makers were more lenient policy choices would change the syst overall outcomes but also imply a higher value of the enhanced surveillance system.

@@

NPI effectiveness (i.e., the % reduction in infectious contacts caused by NPIs) had a non-monotonic effect on the value of surveillance systems to pandemic response (**Supplementary Figure X**). If NPI effectiveness is 0%, an improved surveillance system has no value, as expected. The value of the enhanced surveillance system increases with NPI effectiveness slowly for maximum NPI effectiveness below around 50%, and peaked when the maximum NPI effectiveness reaches 60%, enough to reduce the baseline R0 of 2.5 below 1 at pandemic outset. In general, we find that the value of the EWS peaks at a maximum NPI effectiveness value at $1-1/R0$, for different values of R0. Beyond that point, increases in NPI effectiveness result in lower total pandemic costs, but the value of the surveillance system starts to decrease.

<!--

# If we have time to look at those other cases:

## EWS and policy coordination
EWS alleviate lack of policy coordination. 
There is a concern that EWS data will need to be integrated to be useful. However, we find that policy coordination and EWS act as substitutes.

## If we have time: Marginal Value of EWS
EWS deployment also face decreasing marginal returns. The value of EWS decreases as coverage increases.

-->

>>>>

---
title: "results-draft"
author: "Pedro Nascimento de Lima"
date: "2024-03-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

@@

# Population
The model simulates disease percolating through six interconnected jurisdictions that independently make nonpharmaceutical decisions. Individuals can travel and work across jurisdiction lines, and the six jurisdictions each have a population of 10^6 individuals. Mortality parameters were chosen to reflect mortality rates observed in the United States during the first year of the COVID-19 pandemic; hence our results can be interpreted as expected outcomes for a cluster of US counties.

@@

# Calibration: Add as a footnote to tau.
We use parameter estimates from the literature whenever possible (Supplementary Table 1). Parameters that are unobserved in the data, notably tau and. R0 is computed based on disease progression rates. We calibrate only the $tau$ parameter, which is highly heterogeneous and context-dependent (cite Laura), to match the number of deaths observed during the first year in the United States. We find that $tau = 0.137$. We explore the implication of a wider range of policy effectiveness in sensitivity analyses and explore this particular parameter in more detail.

@@

<!--
Points from this section for discussion:
* **EWS provides meaningful value under base-case assumptions.** Net-monetary benefit from the surveillance system is on the order of 1,000 dollars per person over a one-year time-frame. This outweighs any available estimate of costs.
* **Base-case results align with observed COVID-19 data.** US obseved on the order of 160 deaths per 100,000 people during the first year of the pandemic, and many states.
-->

## Surveillance net benefit depends on epidemiological and societal parameters

@@

A pathogen that is 50% more transmissible than wild-type SARS-Cov-2 would pose a greater challenge for pandemic response, causing `r pull_estimate(r$table_1_long_all, Scenario == "1.5x transmissible", variable == "deaths_per_100k", NMB_comparator == T)` deaths per 100,000 people even if mitigated with the same NPI policy considered in our base-case scenario (Supplementary Table 2). Such a pandemic would cause a substantial societal cost of `r pull_estimate(r$table_1_long_all, Scenario == "1.5x transmissible", variable == "C", NMB_comparator == T)` dollars per person. In those conditions, the EWS would be particularly valuable, reducing the death toll of the pandemic to `r pull_estimate(r$table_1_long_all, Scenario == "1.5x transmissible", variable == "deaths_per_100k", NMB_comparator == F)` per 100,000 people, and would be projected to provide an NMB of `r pull_estimate(r$table_1_long_all, Scenario == "1.5x transmissible", variable == "NMB", NMB_comparator == F)` dollars per person. Conversely, a pathogen that is 50% less transmissible than SARS-Cov-2 would result in a lower death toll of `r pull_estimate(r$table_1_long_all, Scenario == "0.5x transmissible", variable == "deaths_per_100k", NMB_comparator == T)` without the EWS system, reducing the net-monetary benefit of the system to `r pull_estimate(r$table_1_long_all, Scenario == "0.5x transmissible", variable == "NMB", NMB_comparator == F)` dollars per person. The value of the EWS system also depends on the infection fatality rate of the pathogen, but the net benefit of the EWS would be positive even for a pathogen that is half as deadly as SARS-Cov-2. All else equal, the EWS would still yield an NMB of `r pull_estimate(r$table_1_long_all, Scenario == "0.5x deadly", variable == "NMB", NMB_comparator == F)` dollars per person in a pandemic caused by a pathogen that is 50% less deadly than SARS-Cov-2, and would yield `r pull_estimate(r$table_1_long_all, Scenario == "1.5x deadly", variable == "NMB", NMB_comparator == F)` dollars per person in responding to a pathogen that is 50% more deadly than COVID-19.

@@

Longer decision and implementation lags have an overall negative effect on the total costs of the pandemic, longer decision lags also would *increase* the net benefit of EWS systems for pandemic response. Our base-case scenario assumes that decision-makers take one week to decide and implement non-pharmaceutical interventions. If policymakers took two weeks to introduce and implement interventions, the total pandemic cost increases from `r pull_estimate(r$table_1_long_all, Scenario == "Base-case", variable == "C", NMB_comparator == T)` to `r pull_estimate(r$table_1_long_all, Scenario == "2-week decision lag", variable == "C", NMB_comparator == T)` dollars per person, assuming conventional surveillance. The early warning system effectively helps to roll-back decision lags, reducing total costs to `r pull_estimate(r$table_1_long_all, Scenario == "2-week decision lag", variable == "C", NMB_comparator == F)` dollars per person, yielding an NMB of `r pull_estimate(r$table_1_long_all, Scenario == "2-week decision lag", variable == "NMB", NMB_comparator == F)` relative to conventional surveillance. Conversely, in contexts where NPI implementation is swift (i.e, decisions are made within 3 days), the NMB of EWS would be lower but still positive at `r pull_estimate(r$table_1_long_all, Scenario == "3-day decision lag", variable == "NMB", NMB_comparator == F)` dollars per person.

Policy stringency (i.e., how stringent the case thresholds used to introduce or remove interventions) has a similar effect on the value of EWS relative to decision lags - the EWS provides the most value under the least desirable conditions for disease prevention. In a context where decision-makers are 50% more stringent than in our base-case assumptions (i.e., their case thresholds are half of its base-case), society is able to avert a substantial number of deaths, reducing mortality from `r pull_estimate(r$table_1_long_all, Scenario == "Base-case", variable == "deaths_per_100k", NMB_comparator == T)` to `r pull_estimate(r$table_1_long_all, Scenario == "1.5x stringent", variable == "deaths_per_100k", NMB_comparator == T)` deaths per 100,000 population without the EWS. In that scenario, the EWS yields `r pull_estimate(r$table_1_long_all, Scenario == "1.5x stringent", variable == "NMB", NMB_comparator == F)` dollars per person of net benefit since most of the deaths would already be prevented through more stringent NPIs. Conversely, if policymakers were more lenient policy choices would change the system overall outcomes but also imply a higher value of the enhanced surveillance system.

@@

NPI effectiveness (i.e., the % reduction in infectious contacts caused by NPIs) had a non-monotonic effect on the value of surveillance systems to pandemic response (**Supplementary Figure X**). If NPI effectiveness is 0%, an improved surveillance system has no value, as expected. The value of the enhanced surveillance system increases with NPI effectiveness slowly for maximum NPI effectiveness below around 50%, and peaks when the maximum NPI effectiveness reaches 60%, enough to reduce the baseline R0 of 2.5 below 1 at the pandemic onset. In general, we find that the value of the EWS peaks at a maximum NPI effectiveness value at $1-1/R0$, for different values of R0. Beyond that point, increases in NPI effectiveness result in lower total pandemic costs, but the value of the surveillance system starts to decrease.
